---
title: "Project1"
author: Lindsey Marquez
output: word_document
---

```{r setup, include=TRUE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

# Read file
file <- "TCGA_breast_cancer_ERpositive_vs_ERnegative_PAM50.tsv"

first10 <- c('NAT1','BIRC5','BAG1','BCL2','BLVRA','CCNB1','CCNE1','CDC6','CDC20','CDH3')
```
## Assignment

**1.** For the assignment use the second dataset called TCGA_breast_cancer_ERpositive_vs_ERnegative_PAM50.tsv that shows ER assignment for each sample (Positive vs. Negative).
 
**2.** Compute 5-fold and 10-fold cross-validation estimates of prediction accuracies of ER using all genes by utilizing logistic regression and compare with NNC (2x2 table).

**3.** Modify the the R markdown document template to report your computation and results in a table format.

**4.** Comment on the quality of results

**5.** In the second part of the assignment use Project1fs.R to process a large data set by first removing all genes with sd < 1 and subsequently use Feature selection to pick top 50 genes vs top 100 genes for cross-validation based on the t-test statistic.

**6.** For extra credit – please replace centroid based classifier with one utilizing logistic or lasso regression similarly to the first part of the assignment and report on any difficulties.

## Reading data

Please add R code that reads data here - 
reading file: `r file`

```{r reading_data, echo=TRUE}
system.time({
# important -- this makes sure our runs are consistent and reproducible
set.seed(0)

header <- scan(file, nlines = 1, sep="\t", what = character())
data <- read.table(file, skip = 2, header = FALSE, sep = "\t", quote = "", check.names=FALSE)

header[1] <- "gene_id"
names(data) <- header

header2 <- scan(file, skip = 1, nlines = 1, sep="\t", what = character())

Positive <-  data[, header2 == "Positive", drop = FALSE]
Negative <-  data[, header2 == "Negative", drop = FALSE]

})
```

## Computation

Please add R code that computes the results

```{r computation, echo=TRUE}

# Define function cross_valid so we can rerun the cross validation with various parameters

cross_validation <- function (nfold, alg="centroid") {

# Split each cancer type samples into nfold groups
Positive_groups <- split(sample(colnames(Positive)), 1+(seq_along(colnames(Positive)) %% nfold))
Negative_groups <- split(sample(colnames(Negative)), 1+(seq_along(colnames(Negative)) %% nfold))

result <- array()

# Iterate from 1 to nfold groups -- to choose test group
for (test_group in 1:nfold) {
  
  # Return all samples in the chosen test group
  testPositive <- Positive[,colnames(Positive) %in% unlist(Positive_groups[test_group])]
  testNegative <- Negative[,colnames(Negative) %in% unlist(Negative_groups[test_group])]
  
  # Return all samples *not* in the chosen test group 
  trainingPositive <- Positive[,!(colnames(Positive) %in% unlist(Positive_groups[test_group]))]
  trainingNegative <- Negative[,!(colnames(Negative) %in% unlist(Negative_groups[test_group]))]
  
  # Centroid function
  if (alg == "centroid") { 
    
    #Take row means to calculate centroids
    centroidERpos <- rowMeans(trainingPositive) 
    centroidERneg <- rowMeans(trainingNegative) 
    
    # Calculate misclassified centroid predictions 
    misclassifiedERpos <- sum(apply(testPositive, 2, function(x) sum(abs(x - centroidERpos)) > sum(abs(x - centroidERneg))))
    misclassifiedERneg <- sum(apply(testNegative, 2, function(x) sum(abs(x - centroidERpos)) < sum(abs(x - centroidERneg)))) 
  }
  
  # General linear model function
  if (alg == "GLM") { 
    
  # Build training data frame 
  trainingER <- rbind( cbind(data.frame(t(trainingPositive)), cancer = 1),   cbind(data.frame(t(trainingNegative)), cancer = 0) )
  
  # Build test data set 
  test_ER <- rbind( data.frame(t(testPositive)), 
              data.frame(t(testNegative)))
  
  # Provide true labels 
  truth <- c(rep(1, ncol(testPositive)), rep(0, ncol(testNegative)))
  
  # Create logistic regression model
  model <- glm(cancer ~ ., data = trainingER, family = binomial)
  
  # Prediction model using regression
  p <- predict(model, newdata = test_ER, type = "response")
  preds <- ifelse(p < 0.5, 0, 1)
  
  # Calculate false positive and false negative predictions
  misclassifiedERpos <- sum(preds[truth == 1] == 0)
  misclassifiedERneg <- sum(preds[truth == 0] == 1) } 
  
  # Calculate fold-level error rate
  result[test_group] <- (misclassifiedERpos + misclassifiedERneg) / (ncol(testPositive) + ncol(testNegative))
  }
  return(list(mean = mean(result), sd = sd(result)))
 c(mean(result), sd(result))
 paste("mean=",mean(result),"sd=",sd(result))
}
system.time({

# 5-fold cross validation
kNNC_5_all <- cross_validation(nfold=5)
GLM_5_all <- cross_validation(nfold=5, alg="GLM")

# 10-fold cross validation
kNNC_10_all <- cross_validation(nfold=10)
GLM_10_all <- cross_validation(nfold=10, alg="GLM")

})

```
## Results

These are our results:

### 5- and 10-fold cross validations
```{r results5, echo=TRUE}

# pull values from each cross validation
cv_meantable <- data.frame(
  GLM = c(GLM_5_all$mean, GLM_10_all$mean),
  kNNC = c(kNNC_5_all$mean, kNNC_10_all$mean)
)

cv_sdtable <- data.frame(
  GLM = c(GLM_5_all$sd, GLM_10_all$sd),
  kNNC = c(kNNC_5_all$sd, kNNC_10_all$sd)
)
# name rows in table
rownames(cv_meantable) <- c("**5-fold**", "**10-fold**")
rownames(cv_sdtable) <- c("**5-fold**", "**10-fold**")

```

### Mean values
```{r mean results, echo=TRUE}
# display 2x2 table
kable(cv_meantable, digits = 4)
```

### SD values
```{r sd results, echo=TRUE}
# display 2x2 table
kable(cv_sdtable, digits = 4)
```

## Discussion

Both cross validations (kNNC and GLM) had misclassifications rates of approximately 7%. The lowest missclassification rate was the 10-fold kNNC with a rate of 6.34%. Based on the values in the mean table above the 10-fold kNNC is suggested to make slightly better predictions than the 5-fold kNNC model.However this trend is opposite when evaluating the GLM results. The 5-fold GLM validation misclassified at 6.76%, while the 10-fold GLM validation had a higher rate of misclassfication at 6.94%.Therefore the GLM values suggest that the 5-fold validation group makes more accurate predictions.

In terms of standard deviation the 5-fold kNNC group had the lowest sd (0.0111) indicating that the data has a smaller spread than the other groups. The remaining groups showed sd values close to 0.02. 

Both algorithms appear to create decent quality predictions due to accuracy values being >90%.

# Part 2

```{r part2, echo=TRUE}
source("Project1fs.R")
```
## Part 2 Results 

```{r part2 results, echo=TRUE}
centroid_res <- cross_validation(5, alg="centroid")
GLM_res <- cross_validation(5, alg="GLM")

comparison <- data.frame(
  Algorithm = c("Centroid.fs", "GLM.fs"),
  Mean = c(centroid_res$mean, GLM_res$mean),
  SD = c(centroid_res$sd, GLM_res$sd)
)

knitr::kable(comparison)
```

## Part 2 Discussion

When using feature selection on this dataset the centroid based model had a misclassification rate of 6.2% while the GLM model had a higher misclassification rate which was 7.8%.These values suggest that the centroid algorithm has the potential to make more accurate predictions than the regression based algorithm. 

The spread of sd for the feature selection groups showed that the centroid model had a small spread with an sd of 0.0391 while the regression model had an sd of 0.054. This is similar what was observed in the models from part 1 where we observed that the centroid group had the lowest spread based on sd value.

Both algorithms used in feature selection are have accuracy values >90%, which indicates good performance from our model.

